Deep mutational scanning of hemagglutinin helps identify evolutionarily successful human H3N2 influenza viruses

### 2018-01-30 paper/manuscript.tex
added some of John edits

\title{Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages}

@@

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid constraints measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that the deep mutational scanning captures much of the purifying selection on HA.
ExpCM fit a stringency parameter that relates the selection in the experiments to that in nature~\cite{bloom2017identification,hilton2017phydms}.
The stringency parameter for our HA ExpCM is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

@@

\subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
The results above show that measurements of mutational effects in the Perth/2009 HA are informative about the evolutionary fate of other human H3N2 HAs.
How broadly can such measurements be generalized across HAs?
To address this question, we repeated the trunk versus side-branch analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
Figure~\ref{fig:WSN_trunkvssidebranch} shows that the H1 measurements are not informative for distinguishing the trunk and side branches of an H3 phylogeny.
This fact indicates that the utility of an experiment for discriminating successful and unsuccessful strains degrades over sufficiently long evolutionary distances.

@@

Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}.
One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}A).
Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
The clade containing H1 has an upward shift of the globular head relative to the clade containing H3.
This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.

>>>>

\title{Deep mutational scanning of hemagglutinin helps identify evolutionarily successful human H3N2 influenza viruses}

@@

\subsection*{Our measurements are consistent with existing knowledge about HA's evolution and function}
How do the HA amino-acid preferences measured in our experiments relate to the evolution of H3N2 influenza virus in nature?
This question can be addressed by evaluating how well an experimentally informed codon substitution model (ExpCM) using our measurements describes H3N2 evolution compared to standard phylogenetic substitution models~\cite{bloom2017identification,hilton2017phydms}.
Table~\ref{tab:phydms} shows that the ExpCM using the across-replicate average of our measurements greatly outperforms conventional substitution models.
This result indicates that our experiments authentically capture some of the constraints on HA evolution. 
The relative rate of nonsynonymous to synonymous substitutions (dN/dS or $\omega$) is $\ll$1 for conventional substitution models (Table~\ref{tab:phydms}).
However, the relative rate of nonsynonymous to synonymous substitutions after accounting for the amino-acid preferences measured in our experiments (the $\omega$ for the ExpCM) is close to one (Table~\ref{tab:phydms}), indicating that most purifying selection against nonsynonymous substitutions is accounted for by the constraints measured in the deep mutational scanning.
The ExpCM stringency parameter~\cite{hilton2017phydms} is 2.47 (Table~\ref{tab:phydms}), indicating that natural selection favors the same amino acids as the experiments but with greater stringency.
Throughout the rest of this paper, we use experimental measurements re-scaled~\cite{bloom2017identification,hilton2017phydms} by this stringency parameter.
These re-scaled preferences are shown in Figure~\ref{fig:logoplot}.

@@

\subsection*{There are large differences between H3 and H1 HAs in the amino-acid preferences of many sites}
The results above show that measurements of mutational effects in the Perth/2009 HA are informative about the evolutionary fate of other human H3N2 HAs.
To determine how broadly experimental measurements be generalized across HAs, we repeated the trunk versus side-branch analyses using amino-acid preferences from our prior deep mutational scanning of the WSN/1933 H1 HA, which is highly diverged from the Perth/2009 H3 HA.
Figure~\ref{fig:WSN_trunkvssidebranch} shows that the H1 measurements are not informative for distinguishing the trunk and side branches of an H3 phylogeny.
This fact indicates that the utility of an experiment for discriminating successful and unsuccessful strains degrades over sufficiently long evolutionary distances.

@@

Despite their high sequence divergence, H1 and H3 adopt very similar protein folds~\cite{ha2002h5,russell2004h1}.
However, there are differences in the rotation and upward translation of the globular head subdomains relative to the central stalk domain among different HA subtypes~\cite{ha2002h5,russell2004h1}.
Previous work has defined clades of structurally related HA subtypes~\cite{ha2002h5,russell2004h1}.
One such clade includes H1, H2, H5, and H6, whereas another clade includes H3, H4, and H14 HAs (Figure~\ref{fig:distance_distribution}A).
Sites that are conserved at different amino-acid identities in these two clades tend to have exceptionally large shifts in amino acid preferences (Figure~\ref{fig:RMSD_heatmap}B).
The clade containing H1 has an upward translation of the globular head relative to the clade containing H3.
This structural shift has been attributed largely to the interaction between sites 107 and 75(HA2)~\cite{ha2002h5,russell2004h1}.
Specifically, the clade containing H1 has a taller turn in the interhelical loop connecting helix A and helix B in the stalk domain, and this tall turn is stabilized by a hydrogen bond between Glu-107 and Lys-75(HA2) (Figure~\ref{fig:RMSD_heatmap}C).
In deep mutational scanning of the H1 HA, site 107 has a high preference for Glu and 75(HA2) strongly prefers positively charged Lys and Arg.
In contrast, the interhelical loop in H3 HA makes a sharper and shorter turn which is facilitated by a Gly at 75(HA2).
In the deep mutational scanning of the Perth/2009 H3 HA, site 75(HA2) prefers Gly and to a lesser extent Val, while site 107 is fairly tolerant of mutations.
Therefore, some of the shifts in HA amino-acid preferences can be directly rationalized in terms of changes in HA structure.
